---
title: "Phase 2 — Exploratory Data Analysis"
subtitle: "CTA System-wide Daily Ridership (bus / rail / total)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 9, fig.height = 4.5)

library(tidyverse)
library(tsibble)
library(feasts)
library(lubridate)

# Run src/data_prep.py first so this file exists.
clean_path <- here::here("data", "processed", "clean_v1.csv")
stopifnot(file.exists(clean_path))

df <- read_csv(clean_path, show_col_types = FALSE) %>%
  mutate(service_date = as_date(service_date))
```

## 1. Load as a tsibble

Each `mode` (bus / rail / total) is its own key — this keeps the three
series distinct throughout `feasts`/`fable` without duplicating code per
mode.

```{r as-tsibble}
rides_tsbl <- df %>%
  as_tsibble(index = service_date, key = mode) %>%
  fill_gaps()

rides_tsbl
```

## 2. Overall trend over time

```{r trend-plot}
rides_tsbl %>%
  ggplot(aes(x = service_date, y = rides, color = mode)) +
  geom_line(alpha = 0.6) +
  labs(title = "CTA daily ridership, 2001–present", x = NULL, y = "Rides",
       color = "Mode") +
  theme_minimal()
```

**Look for:** the pre-2020 level, the COVID-19 cliff in spring 2020, and
whether ridership has since recovered, plateaued, or shifted composition
between bus and rail. This is the kind of structural break the brief asks
you to name explicitly.

## 3. Day-of-week seasonality

```{r dow-seasonality}
df %>%
  mutate(day_of_week = wday(service_date, label = TRUE, week_start = 1)) %>%
  ggplot(aes(x = day_of_week, y = rides, fill = mode)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~mode, scales = "free_y") +
  labs(title = "Ridership by day of week", x = NULL, y = "Rides") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 4. Holiday / weekend effects

The raw `day_type` field (W/A/U) already encodes weekday vs. Saturday vs.
Sunday-or-holiday. Compare distributions directly:

```{r daytype-effect}
df %>%
  mutate(day_type = recode(day_type, W = "Weekday", A = "Saturday", U = "Sunday/Holiday")) %>%
  ggplot(aes(x = day_type, y = rides, fill = mode)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~mode, scales = "free_y") +
  labs(title = "Ridership by CTA day type", x = NULL, y = "Rides") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 5. Outlier days flagged in Phase 1

`data_prep.py` flags statistically extreme days with a rolling robust
z-score (documented there — flagged, not dropped). Check whether they
cluster around known events rather than looking like noise:

```{r outlier-check}
df %>%
  filter(is_outlier, mode == "total") %>%
  arrange(service_date) %>%
  select(service_date, day_type, rides) %>%
  head(20)
```

## 6. Stationarity: ACF / PACF and a unit-root test

Using the `total` series as the representative one (bus and rail move
together closely enough that checking all three rarely changes the
conclusion — note in your write-up if that's not the case for your run).

```{r acf-pacf}
total_tsbl <- rides_tsbl %>% filter(mode == "total")

total_tsbl %>%
  ACF(rides, lag_max = 28) %>%
  autoplot() + labs(title = "ACF — total ridership")

total_tsbl %>%
  PACF(rides, lag_max = 28) %>%
  autoplot() + labs(title = "PACF — total ridership")
```

```{r unit-root}
total_tsbl %>%
  features(rides, unitroot_kpss)

total_tsbl %>%
  features(rides, unitroot_ndiffs)
```

`unitroot_kpss` p-value below 0.05 suggests the series is NOT
stationary (rejects the null of stationarity); `unitroot_ndiffs` gives the
number of first differences KPSS suggests you'd need to reach stationarity.

## 7. Interpreting the diagnostics

Use the plots and test output above to describe the behavior in the data
snapshot you loaded. Check whether weekly patterns are visible, how
ridership changes around 2020 and afterward, what the KPSS and differencing
tests suggest, and whether flagged outliers align with dates worth
investigating. These diagnostics describe this dataset; they do not
establish that any one forecasting model will perform best. Compare model
choices against the held-out results from
`notebooks/03_baseline_models.Rmd` and the Python training pipeline.
