---
title: "Phase 4 — Baseline Models"
subtitle: "Naive / Seasonal Naive / ETS / ARIMA / Prophet, per mode"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(tsibble)
library(fable)
library(fable.prophet)
library(fabletools)
library(lubridate)

clean_path <- here::here("data", "processed", "clean_v1.csv")
stopifnot(file.exists(clean_path))

df <- read_csv(clean_path, show_col_types = FALSE) %>%
  mutate(service_date = as_date(service_date))

rides_tsbl <- df %>%
  as_tsibble(index = service_date, key = mode) %>%
  fill_gaps()
```

## 1. Reproduce the SAME held-out window Python uses

`src/features.py::time_based_split` defines `test_start = max_date -
TEST_WEEKS + 1 day` with `TEST_WEEKS = 10`, `VAL_WEEKS = 6`. We replicate
that arithmetic here (not the val cut — baselines like ETS/ARIMA/Prophet
select their own structure via AICc/internal CV rather than needing a
separate validation set the way Phase 5's hyperparameter search does, so
they're trained on everything before the test window).

```{r split}
TEST_WEEKS <- 10

max_date <- max(rides_tsbl$service_date, na.rm = TRUE)
test_start <- max_date - weeks(TEST_WEEKS) + days(1)

train_tsbl <- rides_tsbl %>% filter(service_date < test_start)
test_tsbl  <- rides_tsbl %>% filter(service_date >= test_start)

cat("Test window:", format(test_start), "to", format(max_date), "\n")
```

## 2. Fit all five baselines, per mode, in one mable

```{r fit}
fit_time <- system.time({
  fit <- train_tsbl %>%
    model(
      naive  = NAIVE(rides),
      snaive = SNAIVE(rides ~ lag("week")),
      ets    = ETS(rides),
      arima  = ARIMA(rides),
      # Weekly + annual Fourier terms via Prophet's own seasonality args.
      # If Phase 2's EDA didn't show a visible annual pattern for a given
      # mode, this term will just come out near-zero — harmless to leave in.
      prophet = prophet(rides ~ season(period = 7, order = 3) +
                                 season(period = 365.25, order = 6))
    )
})["elapsed"]

fit
```

## 3. Forecast into the test window and score

```{r forecast-and-score}
fc_time <- system.time({
  fc <- fit %>% forecast(new_data = test_tsbl)
})["elapsed"]

fc_tbl <- fc %>%
  as_tibble() %>%
  select(mode, .model, service_date, .mean)

actual_tbl <- test_tsbl %>%
  as_tibble() %>%
  select(mode, service_date, rides)

joined <- fc_tbl %>% left_join(actual_tbl, by = c("mode", "service_date"))

# NOTE: fit_time / fc_time are measured for the WHOLE batch of 5 models
# together (one model() / forecast() call fits all keys x all formulas at
# once) rather than per individual model — separating that out isn't the
# point of the baseline phase. Take these as "cost to produce all five
# baselines", not a per-model number; Phase 5/6 report true per-model times.
metrics <- joined %>%
  group_by(mode, .model) %>%
  summarise(
    MAE   = mean(abs(rides - .mean)),
    RMSE  = sqrt(mean((rides - .mean) ^ 2)),
    MAPE  = mean(abs((rides - .mean) / rides)) * 100,
    sMAPE = mean(2 * abs(rides - .mean) / (abs(rides) + abs(.mean))) * 100,
    .groups = "drop"
  ) %>%
  rename(model = .model) %>%
  mutate(
    training_time_s   = round(as.numeric(fit_time), 3),
    inference_time_s  = round(as.numeric(fc_time), 4),
    interpretability  = case_when(
      model == "naive"   ~ "high (trivial, no parameters)",
      model == "snaive"  ~ "high (trivial, seasonal lag only)",
      model == "ets"     ~ "high (well-understood exponential-smoothing components)",
      model == "arima"   ~ "high (interpretable AR/MA/seasonal terms)",
      model == "prophet" ~ "medium (interpretable trend/seasonality decomposition)",
      TRUE ~ NA_character_
    )
  )

metrics %>% arrange(mode, RMSE) %>% knitr::kable(digits = 2)
```

This is the "line in the sand" table — every model in Phases 5-6 needs to
beat the best row here per mode to justify its added complexity.

## 4. Save for Phase 7 consolidation

```{r save}
out_path <- here::here("reports", "baseline_metrics.csv")
dir.create(dirname(out_path), showWarnings = FALSE, recursive = TRUE)
write_csv(metrics, out_path)
cat("Saved ->", out_path, "\n")
```

## 5. Residual diagnostics (pick your best baseline per mode)

```{r residuals}
fit %>%
  select(mode, ets) %>%
  filter(mode == "total") %>%
  gg_tsresiduals() +
  labs(title = "ETS residual diagnostics — total ridership")
```

Look for: residual autocorrelation left in the ACF (suggests the model
hasn't captured all the structure), and whether residuals look roughly
normal (Ljung-Box test via `fit %>% augment() %>% features(.innov,
ljung_box, lag = 14)` if you want a formal test to report).
